---
title: "My Dataset Search"
output: html_document
date: "2026-01-26"
---

# Setting up `GEOmetadb`

`GEOmetadb` provides a SQLite mirror of GEO metadata that you can query from R.
It mirrors GEO as-is (no curation). Download the current SQLite file once and reuse it.

## Install packages

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
if (!requireNamespace("GEOmetadb", quietly = TRUE))
  BiocManager::install("GEOmetadb")
if (!requireNamespace("DBI", quietly = TRUE))
  install.packages("DBI")
if (!requireNamespace("RSQLite", quietly = TRUE))
  install.packages("RSQLite")
if (!requireNamespace("dplyr", quietly = TRUE))
  install.packages("dplyr")
if (!requireNamespace("ggplot2", quietly = TRUE))
  install.packages("ggplot2")
```

## Get the SQLite database

```{r}
library(GEOmetadb)
if (!file.exists('GEOmetadb.sqlite')) getSQLiteFile()
file.info('GEOmetadb.sqlite')
```

## Connect to the database

```{r}
library(DBI)
library(RSQLite)
con <- dbConnect(SQLite(), 'GEOmetadb.sqlite')
```

## Explore tables and fields

```{r}
dbListTables(con)
dbListFields(con, 'gse')
```


# Exploring Platforms (GPL)

How many platforms exist, and what technologies do they use?

```{r}
num_platforms <- dbGetQuery(con, 'select count(*) as n from gpl')
num_platforms

uniq_tech <- dbGetQuery(con, 'select distinct technology from gpl')
NROW(uniq_tech)

num_uniq_tech <- dbGetQuery(con, 'select technology, count(*) as Num_Platforms from gpl group by technology')
# Quick plot
library(ggplot2)
plot_df <- subset(num_uniq_tech, !is.na(technology))
 ggplot(plot_df, aes(x = technology, y = Num_Platforms)) +
   geom_col() + coord_flip()
```

## Restrict to human

```{r}
num_uniq_tech_human <- dbGetQuery(con, "select technology, count(*) as Num_Platforms from gpl where organism = 'Homo sapiens' group by technology order by Num_Platforms desc")
num_uniq_tech_contain_human <- dbGetQuery(con, "select technology, count(*) as Num_Platforms from gpl where organism like '%Homo sapiens%' group by technology order by Num_Platforms desc")

library(dplyr)
uniq_tech_all <- bind_rows(
  transform(num_uniq_tech_human, type = 'human only'),
  transform(num_uniq_tech_contain_human, type = 'contains human'),
  transform(num_uniq_tech, type = 'All species')
)

ggplot(uniq_tech_all, aes(technology, Num_Platforms, fill = type)) +
  geom_col(position = 'dodge') + coord_flip() +
  theme(legend.position = 'bottom') +
  ggtitle('Platforms that contain human vs human only')
```


# Finding Datasets of Interest

Goal: RNA-seq, human, recent (last ~3 years), and with sufficient samples.

## RNA-seq series by platform technology

```{r}
sql <- paste(
  "SELECT DISTINCT gse.title, gse.gse, gpl.title",
  "FROM gse JOIN gse_gpl ON gse_gpl.gse = gse.gse",
  "JOIN gpl ON gse_gpl.gpl = gpl.gpl",
  "WHERE gpl.technology = 'high-throughput sequencing'"
)
rs <- dbGetQuery(con, sql)
dim(rs)
```

## Add organism and recency filters

```{r}
sql <- paste(
  "SELECT DISTINCT gse.title, gse.gse, gpl.title, gse.submission_date",
  "FROM gse JOIN gse_gpl ON gse_gpl.gse = gse.gse",
  "JOIN gpl ON gse_gpl.gpl = gpl.gpl",
  "WHERE gse.submission_date > '2022-01-01' AND",
  "gpl.organism LIKE '%Homo sapiens%' AND",
  "gpl.technology LIKE '%high-throughput seq%'"
)
rs <- dbGetQuery(con, sql)
dim(rs)
```

## Add a topic keyword (e.g., ovarian)

```{r}
sql <- paste(
  "SELECT DISTINCT gse.title, gse.gse, gpl.title, gse.submission_date",
  "FROM gse JOIN gse_gpl ON gse_gpl.gse = gse.gse",
  "JOIN gpl ON gse_gpl.gpl = gpl.gpl",
  "WHERE gse.submission_date > '2022-01-01' AND",
  "gse.title LIKE '%Sarcoma%' AND",
  "gpl.organism LIKE '%Homo sapiens%' AND",
  "gpl.technology LIKE '%high-throughput seq%'"
)
rs <- dbGetQuery(con, sql)
dim(rs)
head(rs[, c('gse','title')])
```

## Heuristic: require a counts-like supplementary file

```{r}
sql <- paste(
  "SELECT DISTINCT gse.title, gse.gse, gpl.title, gse.submission_date, gse.supplementary_file",
  "FROM gse JOIN gse_gpl ON gse_gpl.gse = gse.gse",
  "JOIN gpl ON gse_gpl.gpl = gpl.gpl",
  "WHERE gse.submission_date > '2022-01-01' AND",
  "gse.title LIKE '%Sarcoma%' AND",
  "gpl.organism LIKE '%Homo sapiens%' AND",
  "gpl.technology LIKE '%high-throughput sequencing%'",
  "ORDER BY gse.submission_date DESC"
)
rs <- dbGetQuery(con, sql)
counts_files <- rs$supplementary_file[grep("count|cnt", rs$supplementary_file, ignore.case = TRUE)]
series_of_interest <- rs$gse[grep("count|cnt", rs$supplementary_file, ignore.case = TRUE)]
# Preview filenames
shortened <- unlist(lapply(counts_files, function(x){
  x <- unlist(strsplit(x, ";"))
  x <- x[grep("count|cnt|txt|csv|xlsx|TAR", x, ignore.case = TRUE)] # included TAR in file name
  tail(unlist(strsplit(x, "/")), n = 1)
}))
head(shortened, 10)
```

## Require a minimum number of samples per series

```{r}
num_series <- dbGetQuery(con, paste(
  "select * from gsm where series_id in ('",
  paste(series_of_interest, collapse = "','"), "')", sep = ""))

gse.count <- as.data.frame(table(num_series$series_id))
subset(gse.count, Freq > 6)
```

## Selected GSE198568: Full-length transcriptome sequencing reveals gene expression signatures in human soft tissue sarcomas